Approximation as Differentiation

### 2025-06-09 main/first-order.tex
Onto AutoDiff.

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} monad $\synVar{L}$. Rather than the usual definition of
the type of lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then define
(within the language) a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat}
\times \alpha)$, along with an approximating version of multiplication $*_{\text{appr}}:
\synVar{L}\,\synVar{nat} \times \synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$ given as as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' =
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\].

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$ in
\exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking this specific approximating behaviour into the interpretation of
individual operators.

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic ideas here.

\subsubsection{Derivatives of Differentiable Functions}

For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
\RR^n$ at a point $x$ have the following form \todo{Can probably delete now as covered in
\secref{approx-as-tangents}}:

\begin{itemize}
\item The \emph{forward derivative} (tangent map or pushforward) $\pushf{f}_x$ of $f$ at $x \in \RR^m$ is the
unique linear map $\RR^m \linearto \RR^n$ given by the Jacobean matrix of $f$ at $x$.
\item The \emph{backward derivative} (cotangent map or pullback) $\pullf{f}_x$ is the unique linear map
$\RR^n \linearto \RR^m$ given by the transpose (adjoint) of the Jacobean matrix of $f$ at $x$.
\end{itemize}

@@

Derivatives respect composition~\cite{spivak65}. Suppose $f: \RR^m \to \RR^n$ and $g: \RR^n \to \RR^k$. Then
for any $x \in \RR^m$ we have:

@@

We can verify that $\tangents$ is functorial, regardless of whether the linear maps $\RR^m \linearto \RR^n$
happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the
forward derivative of a composite function, without have to recompute the function itself, as long as all
atomic operations provide derivatives. An efficient implementation of \GPS based on $\Fam(\LatGal)$
(\secref{fam:galois-slicing} above) would most likely apply a similar optimisation to make it possible to
compute forwards slices without having to reexecute the program at the same time.

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the constant family of tangent spaces at $\RR^n$ (also written $\RR^n$,
for convenience), and each map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as the pair $(f_1, f_2)$.
Then the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f)$ again corresponds exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp
\partial f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$, with the reindexing along $f$ recomputing $f(x)$ in
order to select the appropriate pushforward.

>>>>

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} monad $\synVar{L}$. Rather than the usual definition of
the type of lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then define
(within the language) a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat}
\times \alpha)$, along with an approximating version of multiplication $*_{\text{appr}}:
\synVar{L}\,\synVar{nat} \times \synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$ expressible as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' =
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\].

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$
from \exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking this specific approximating behaviour into the interpretation of
individual operators.

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. First we recall the chain rule for the forward derivative
$\pushf{f}_x$ and backward derivative $\pullf{f}_x$, which says that derivatives respect
composition~\cite{spivak65}.

% \subsubsection{Derivatives of Differentiable Functions}
%
% For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
% isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
% \RR^n$ at a point $x$ have the following form \todo{Can probably delete now as covered in
% \secref{approx-as-tangents}}:
%
% \begin{itemize}
% \item The \emph{forward derivative} (tangent map or pushforward) $\pushf{f}_x$ of $f$ at $x \in \RR^m$ is the
% unique linear map $\RR^m \linearto \RR^n$ given by the Jacobean matrix of $f$ at $x$.
% \item The \emph{backward derivative} (cotangent map or pullback) $\pullf{f}_x$ is the unique linear map
% $\RR^n \linearto \RR^m$ given by the transpose (adjoint) of the Jacobean matrix of $f$ at $x$.
% \end{itemize}

@@

Suppose $f: \RR^m \to \RR^n$ and $g: \RR^n \to \RR^k$. Then
for any $x \in \RR^m$ we have:

@@

We can verify that $\tangents$ is functorial, regardless of whether the linear maps $\RR^m \linearto \RR^n$
happen to be derivatives. But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and
similarly for $g$ then the derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is
the derivative of $(g \comp f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the
forward derivative of a composite function, without have to recompute the function itself, as long as all
atomic operations provide derivatives.

% An efficient implementation of \GPS based on $\Fam(\LatGal)$
% (\secref{fam:galois-slicing} above) would most likely apply a similar optimisation to make it possible to
% compute forwards slices without having to reexecute the program at the same time.

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the constant family of tangent spaces at $\RR^n$ (also written $\RR^n$,
for convenience), and each map $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as the pair $(f_1, f_2)$.
Then the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp
\partial f)$ again corresponds exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp
\partial f)_x = {\partial g}_{f(x)} \comp {\partial f}_x$

% with the reindexing along $f$ recomputing $f(x)$ in order to select the appropriate pushforward.
